---
title: "March Madness 2018 EDA"
author: "Jacob Dym, Paul Harmon"
date: "February 23, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Data files
```{r}
# Note this wd setup work only if the current working directory is the MARCHMADNESS folder
#Paul: If on P$'s computer, don't run the next line of code
work <- getwd()
n <- nchar(work)
work <- substr(work, 1, n - 13)
setwd(paste0(work, "/DataFiles"))

# Load DataFiles
Cities <- read.csv("Cities.csv")
Conferences <- read.csv("Conferences.csv")
ConferenceTourneyGames <- read.csv("ConferenceTourneyGames.csv")
GameCities <- read.csv("GameCities.csv")
NCAATourneyCompactReults <- read.csv("NCAATourneyCompactResults.csv")
NCAATourneyDetailedResults <- read.csv("NCAATourneyDetailedResults.csv")
NCAATourneySeedRoundSlots <- read.csv("NCAATourneySeedRoundSlots.csv")
NCAATourneySeeds <- read.csv("NCAATourneySeeds.csv")
NCAATourneySlots <- read.csv("NCAATourneySlots.csv")
RegularSeasonCompactResults <- read.csv("RegularSeasonCompactResults.csv")
RegularSeasonDetailedResults <- read.csv("RegularSeasonDetailedResults.csv")
Seasons <- read.csv("Seasons.csv")
SecondaryTourneyCompactResults <- read.csv("SecondaryTourneyCompactResults.csv")
SecondaryTourneyTeams <- read.csv("SecondaryTourneyTeams.csv")
TeamCoaches <- read.csv("TeamCoaches.csv")
TeamConferences <- read.csv("TeamConferences.csv")
Teams <- read.csv("Teams.csv")
TeamSpellings <- read.csv("TeamSpellings.csv")
```

# EDA
```{r}
library(ggplot2)
library(dplyr)
# Add descriptions to ConferencesTourneyGames
Conferences <- left_join(ConferenceTourneyGames, Conferences, "ConfAbbrev")
rm(ConferenceTourneyGames)

# Look at 2017 NCAA tournament
tournament_2017 <- NCAATourneyCompactReults[NCAATourneyCompactReults$Season == 2017, ]
tournament_2017$WTeamID <- as.factor(tournament_2017$WTeamID)
dim(tournament_2017) # Verify there are 67 games
ggplot(tournament_2017) + geom_bar(aes(WTeamID), fill = "orange") + theme_bw() # Verify the number of wins (6-7 for first place)

# Look at teams proportion of wins in the regular season and if they make it to the tournament
RegularSeasonSub <- list()
for(i in 1985:2017){
  RegularSeasonSub[[i - 1984]] <- subset(RegularSeasonCompactResults, Season == i)
}

# Look at 2017 season
teams2017 <- unique(c(RegularSeasonSub[[33]]$WTeamID, RegularSeasonSub[[33]]$LTeamID))
length(teams2017) # Number of teams in 2017
NumWins2017 <- NumLoss2017 <- InTourney <- c()
for(i in 1:length(teams2017)){
  NumWins2017[i] <- sum(RegularSeasonSub[[33]]$WTeamID == teams2017[i])
  NumLoss2017[i] <- sum(RegularSeasonSub[[33]]$LTeamID == teams2017[i])
  InTourney[i] <- teams2017[i] %in% tournament_2017$WTeamID | teams2017[i] %in% tournament_2017$LTeamID
}
WinProp2017 <- data.frame(Team = teams2017, Wins = NumWins2017 / (NumWins2017 + NumLoss2017), InTournament = InTourney)

# Plot win proportions
ggplot(WinProp2017) + geom_bar(aes(Team, Wins, fill = InTournament), stat = "identity") + theme_bw()
```
It looks like teams with the highest win rate aren't always in the tournament. This is expected because the teams that get into the tournament are either from at-large bids or they win their conference tournament. (Teams with lower win counts probably got in by winning a conference tournament. )

# Set up for ELO
```{r}
# Construct data frames for individual teams to be used for ELO calculations
Team_ELO <- list()
for(i in 1:dim(Teams)[1]){
  dat <- RegularSeasonCompactResults[RegularSeasonCompactResults$WTeamID == Teams$TeamID[i] |
                                       RegularSeasonCompactResults$LTeamID == Teams$TeamID[i], ]
  win <- dat$WTeamID == Teams$TeamID[i]
  season <- dat$Season
  coach <- TeamCoaches[TeamCoaches$TeamID == Teams$TeamID[i], ]
  coach <- coach[coach$FirstDayNum == 0, ]
  dat <- full_join(dat, coach, by = "Season")
  dat <- dat[dat$Season != 2018, ]
  home <- dat$WLoc
  for(j in 1:length(home)){
    if(win[j]){
      home[j] <- home[j]
      next
    }
    if(!win[j]){
      if(home[j] == "N"){
        home[j] <- home[j]
        next
      }
      else home[j] <- "A"
    }
  }
  ot <- dat$NumOT
  game <- data.frame(wt = dat$WTeamID, lt = dat$LTeamID)
  gselect <- game != Teams$TeamID[i]
  opponent <- c()
  for(j in 1:dim(game)[1]){
    opponent[j] <- game[j, gselect[j, ]] 
  }
  opteam <- c()
  for(j in 1:length(opponent)){
    opteam[j] <- as.character(Teams$TeamName[Teams$TeamID == opponent[j]])
  }
  score_difference <- as.numeric(win) * (dat$WScore - dat$LScore) + as.numeric(!win) * (dat$LScore - dat$WScore)
  elo <- c(1000, rep(0, length(win) - 1))
  Team_ELO[[i]] <- data.frame(TeamName = Teams$TeamName[i], TeamID = Teams$TeamID[i], 
                              Season = season, Win = win, PointSpread = score_difference, Location = home,
                              OvertimeRounds = ot, Coach = dat$CoachName, OpponentID = opponent, 
                              Opponent = opteam, Day = dat$DayNum, ELO = elo) 
}
names(Team_ELO) <- Teams$TeamName

# Function to calculate a teams new ELO rating based on a game

Get_ELO <- function(scalar, df, teamid, iter, baseline, I, K){
  if((I == 1985) & (iter == 1)){
    return(baseline)
  }
  if((I > 1985) & (iter == 1)){
    newcheck <- names(ELO_LIST[[I - 1984]][K]) %in% names(ELO_LIST[[I-1985]])
    if(!newcheck){
      return(baseline) 
    }
    else {
      ind <- which(names(ELO_LIST[[I - 1984]][K]) == names(ELO_LIST[[I-1985]]))
      elo <- mean(ELO_LIST[[I-1985]][[ind]])
      return(elo) #I got it to give new D1 teams basline ELOs, however it has a hard time recognizing old vs new D1 teams
    }
  }
  tdat <- df[[which(TEMP_TEAMS == teamid)]][iter, ] #grabs teams previous elo
  Current_ELO <- ELO_LIST[[I - 1984]][[K]][iter - 1]
  opid <- tdat$OpponentID
  odat <- df[[which(TEMP_TEAMS == opid)]]
  pos <- which((odat$Day == tdat$Day) & (odat$TeamID == tdat$OpponentID) & (odat$Season == tdat$Season))
  odat <- odat[pos, ]
  
  Opponent_ELO <- ifelse(pos ==1, baseline, ELO_LIST[[I - 1984]][[which(TEMP_TEAMS == opid)]][pos - 1]) #finds the right row, then uses the previous row's elo score
  w <- as.numeric(tdat$Win)
  dr <- Current_ELO + 37.5 * (tdat$Location == "H") - (Opponent_ELO + 37.5 * (odat$Location == "H"))
  we <- 1 / (10 ^(-dr / 350) + 1) # edited Probability estimation so estimated probabilities had a better range
  elo <- Current_ELO + scalar * abs(tdat$PointSpread) * (w - we)
  if(elo < 1){
    elo <- 1 
  }
  return(elo)
}
```

# Build ELO scores
```{r}
#builds a function to work through the list and calculate team elo
#DEFINE ELO LIST
ELO_LIST <- list()
for (i in 1985:2017){
  ELO_LIST[[i-1984]] <- list() #defines the kth part to be a list, according to Dr. Dym
  szn_filter <- function(df, year){dplyr::filter(df, Season == i)}
  TTEMP_LIST <- lapply(Team_ELO, szn_filter) #generates list of elos for that given year
  TEMP_LIST <- list()
  
  #check if team actually played in that szn
  team_check <- function(TLIST){
    tf <- c()
    for (q in 1:length(TLIST)){
    tf[q] <- TLIST[[q]][1,3] %in% i
      }
  return(tf)}
  tsizzle <- team_check(TTEMP_LIST)
 
  TEMP_LIST <- TTEMP_LIST[tsizzle]
  TEMP_TEAMS <- c()
  for(j in 1:length(TEMP_LIST)){
    TEMP_TEAMS[j] <- TEMP_LIST[[j]]$TeamID[1] 
  }
  
  n_game <- max(unlist(lapply(TEMP_LIST, nrow)))
  n_teams <- length(TEMP_LIST)
  for(j in 1:n_teams){
    ELO_LIST[[i - 1984]][[j]] <- rep(0, nrow(TEMP_LIST[[j]]))
  }
  names(ELO_LIST[[i - 1984]]) <- names(TEMP_LIST)
  for (j in 1:n_game){
    #code here 
    for (k in 1:n_teams){
      if(dim(TEMP_LIST[[k]])[1] < j) next
      # calculate elo for each team 
      ELO_LIST[[i-1984]][[k]][j] <- Get_ELO(scalar = 5, team = TEMP_LIST[[k]]$TeamID[1], df = TEMP_LIST, iter = j, baseline = 5000, I = i, K = k)
    }
    
  }
  #define previous elo as the last one
  
}
```

OK, so we have some Elos... but we're going to have to tune the Elo. Some issues that are coming up is that in the any of the years after the first year, the ELO ratings are starting at 0 rather than at the mean of the previous year. Not sure what's causing that.

Got the ELO scores to not be 0 at the start, function has a hard time recognizing teams that have been in D1 already and just randomly assigns baseline ELO scores to teams. 

IT'S WORKING


#CAN WE PREDICT BASED ON LAST ELO
```{r}
# Put ELO scores back into Team data
for(i in 1:length(Team_ELO)){
  seasons_played <- unique(Team_ELO[[i]]$Season)
  first_season <- min(seasons_played)
  for(j in seasons_played){
    K <- which(names(ELO_LIST[[j - 1984]]) == Team_ELO[[i]]$TeamName[1])
    Team_ELO[[i]][Team_ELO[[i]]$Season == j, ]$ELO <- ELO_LIST[[j - 1984]][[K]]
  }
}

# Basic probability prediction function
# For seasons 2014 - 2017
NCAA_Tourney_sub <- list()
combinations <- list()
for(i in 2014:2017){
  NCAA_Tourney_sub[[i - 2013]] <- subset(NCAATourneyCompactReults, NCAATourneyCompactReults$Season == i)
  combinations[[i - 2013]] <- t(combn(unique(c(NCAA_Tourney_sub[[i - 2013]]$WTeamID, NCAA_Tourney_sub[[i - 2013]]$LTeamID)), 2))
}

# Format ID combinations so lowest one is in first column (submission formatting)
for(i in 1:4){
  for(j in 1:dim(combinations[[i]])[1]){ 
    rmax <- max(combinations[[i]][j, ])
    rmin <- min(combinations[[i]][j, ])
    combinations[[i]][j, ] <- c(rmin, rmax)
  }
}

# Function to retrieve probabilities
Prob_Predict <- function(teams, S){
  name <- matrix(ncol = 2, nrow = dim(teams)[1])
  for(i in 1:dim(teams)[1]){
    name[i, 1] <- as.character(Teams$TeamName[Teams$TeamID == teams[i, 1]]) 
    name[i, 2] <- as.character(Teams$TeamName[Teams$TeamID == teams[i, 2]])
  }
  elos <- matrix(ncol = 2, nrow = dim(teams)[1])
  for(i in 1:dim(teams)[1]){
    elo1t <- Team_ELO[[which(names(Team_ELO) == name[i, 1])]]
    elo1t <- subset(elo1t, elo1t$Season == S)
    elo1t <- elo1t$ELO
    elos[i, 1] <- elo1t[length(elo1t)]
    elo2t <- Team_ELO[[which(names(Team_ELO) == name[i, 2])]]
    elo2t <- subset(elo2t, elo2t$Season == S)
    elo2t <- elo2t$ELO
    elos[i, 2] <- elo2t[length(elo2t)]
  }
  dr <- elos[, 1] - elos[, 2]
  return(1 / (10 ^(-dr / 350) + 1))
}

# Generate predictions for 2014 - 2017 tournaments
Predictions <- list()
for(i in 1:4){
  Predictions[[i]] <- Prob_Predict(combinations[[i]], 2013 + i)
}

# Format submission
subdat <- matrix(ncol = 2, nrow = 4 * length(Predictions[[1]]))
ss <- c("2014", "2015", "2016", "2017")
ids <- matrix(ncol = 4, nrow = length(Predictions[[1]]))
for(i in 1:length(Predictions[[1]])){
  for(j in 1:4){
    d <- combinations[[j]][i, ]
    d <- paste(d, collapse = "_")
    ids[i, j] <- paste(ss[j], d, sep = "_")   
  }
}
idsr <- c(ids[, 1], ids[, 2], ids[, 3], ids[, 4])
props <- c(Predictions[[1]], Predictions[[2]], Predictions[[3]], Predictions[[4]])
subdat <- cbind(idsr, props)
colnames(subdat) <- c("ID", "Pred")
write.csv(subdat, "GrpsSubmission.csv", row.names = FALSE)
```

I got it set so we can test our predictions, we can look at predictions but it's probably easiest to download the results and Kaggle's formula for scoring and just check on our own to tune the model. I made a submission of simple probabilities we placed 5th from the bottom based on the current setup.





